Staged Abstract Interpreters

### 2019-09-07 paper/30_stageconc.tex
typo

\paragraph{Staged Monads}
We use the same monad stack structure as in the unstaged interpreter: a reader
monad with a state monad. The difference here is that, now the monads operate on
staged values.  For brevity, we call them \textit{staged monads}. In the code,
we also use a @Rep@ prefix on the constructors and types to differentiate them.
But it is important to note that instances of the case class @ReaderT@/@StateT@
are not staged, and that the monadic computation, such as functions of @R => M[A]@,
are also not staged. Instead, the internal data that these monads operate on are
staged, i.e., @R@ and @A@ in the reader monad, @S@ and @A@ in the state monad.
The following code snippet shows the idea. We use \hl{light gray} to highlight
where @Rep@ types are added:
\begin{lstlisting}[escapechar=!]
  case class RepReaderT[M[_]: RepMonad, R, A](run: !\hl{Rep[R]}! => M[A]) {
    def flatMap[B](f: !\hl{Rep[A]}! => ReaderT[M, R, B]): RepReaderT[M, R, B] =
      RepReaderT(r => RepMonad[M].flatMap(run(r))(a => f(a).run(r))); ... }
  case class RepStateT[M[_]: RepMonad, S, A](run: !\hl{Rep[S]}! => M[(A, S)]) {
    def flatMap[B](f: !\hl{Rep[A]}! => StateT[M, S, B]): RepStateT[M, S, B] =
      RepStateT(s => RepMonad[M].flatMap(run(s))(as => f(as._1).run(as._2))); ... }
\end{lstlisting}

The function @f@ passed to @flatMap@ is also a function value known at the current stage.
Therefore all invocations of @flatMap@ can be reduced before generating code.
The fact that we only stage data but not monadic computation or monadic
values is the reason that we can peel of the monad stack in the generated code.
Now the type @AnsM[_]@ is instantiated as the same structure as before, but using the
@Rep@ versions:
\begin{lstlisting}
  type R[T] = Rep[T]
  type AnsM[T] = RepReaderT[RepStateT[RepIdM, Store, ?], Env, T]
\end{lstlisting}

@@

In the previous section, we showed the types of @emit_compiled_clo@ and
@emit_ap_clo@ without concrete implementations. In this section, we briefly
discuss the IR nodes generated by them and sketch the code generation.
We define the IR nodes @IRCompiledClo@ and @IRApClo@ using @case class@
extending from @Def[T]@. @Def[T]@ is a built-in type in the LMS framework,
representing next-stage value definitions of type @T@.
\begin{lstlisting}
  case class IRCompiledClo(f: Rep[(ValSt) => ValSt], λ: Lam, ρ: Rep[Env]) extends Def[Value]
  case class IRApClo(rator: Rep[Value], rand: Rep[Value], σ: Rep[Store]) extends Def[ValSt]
\end{lstlisting}

The IR nodes are manipulated by the LMS passes and finally generated as
next-stage Scala code. To generate code, LMS provides an @emitNode@ method for
programmers to control over what code is generated for each kind of nodes. Here,
we match @IRCompiledClo@ and @IRApClo@, and generate definitions via
@emitValDef@ for them.  For compiled closures, we put the function @f@ and its
accompanying syntactic term and environment to their next-stage value
representation @CompiledClo@. For applications, we know the fact that @fun@ is
an instance of @CompiledClo@ at the next stage, and it has a callable field
@f@. Hence, we generate a next-stage function application @f(arg, σ)@.  The
invocations of @quote@ are used to generate proper names for the other
next-stage values.
\begin{lstlisting}
override def emitNode(sym: Sym[Any], rhs: Def[Any]) = rhs match {
  case IRCompiledClo(f, λ, ρ) => emitValDef(sym, s"CompiledClo($\textdollar${quote(f)}, $\textdollar${quote(λ)}, $\textdollar${quote(ρ)})")
  case IRApClo(rator, rand, σ) => emitValDef(sym, s"$\textdollar$rator.f($\textdollar${quote(rand)}, $\textdollar${quote(σ)})")
  ...
}
\end{lstlisting}

>>>>

\paragraph{Staged Monads}
We use the same monad stack structure as in the unstaged interpreter: a reader
monad with a state monad. The difference here is that, now the monads operate on
staged values.  For brevity, we call them \textit{staged monads}. In the code,
we also use a @Rep@ prefix on the constructors and types to differentiate them.
But it is important to note that the instances of @ReaderT@/@StateT@
are not staged, and that the monadic computation, such as functions of @R => M[A]@,
are also not staged. Instead, the internal data that these monads operate on are
staged, i.e., @R@ and @A@ in the reader monad, @S@ and @A@ in the state monad.
The following code snippet shows the idea. We use \hl{light gray} to highlight
where @Rep@ types are added:
\begin{lstlisting}[escapechar=!]
  case class RepReaderT[M[_]: RepMonad, R, A](run: !\hl{Rep[R]}! => M[A]) {
    def flatMap[B](f: !\hl{Rep[A]}! => ReaderT[M, R, B]): RepReaderT[M, R, B] =
      RepReaderT(r => RepMonad[M].flatMap(run(r))(a => f(a).run(r))); ... }
  case class RepStateT[M[_]: RepMonad, S, A](run: !\hl{Rep[S]}! => M[(A, S)]) {
    def flatMap[B](f: !\hl{Rep[A]}! => StateT[M, S, B]): RepStateT[M, S, B] =
      RepStateT(s => RepMonad[M].flatMap(run(s))(as => f(as._1).run(as._2))); ... }
\end{lstlisting}

The function @f@ passed to @flatMap@ is also a value known at the current stage.
Therefore all invocations of @flatMap@ can be reduced before generating code.
The fact that we only stage data but not monadic computation or monadic
values is the reason that we can peel of the monad stack in the generated code.
Now the type @AnsM[_]@ is instantiated as the same structure as before, but using the
@Rep@ versions:
\begin{lstlisting}
  type R[T] = Rep[T]
  type AnsM[T] = RepReaderT[RepStateT[RepIdM, Store, ?], Env, T]
\end{lstlisting}

@@

In the previous section, we showed the types of @emit_compiled_clo@ and
@emit_ap_clo@ without concrete implementations. In this section, we briefly
discuss the IR nodes generated by them and sketch the code generation.
We define the IR nodes @IRCompiledClo@ and @IRApClo@ using @case class@
extending from @Def[T]@. @Def[T]@ is a built-in type in the LMS framework,
representing next-stage value definitions of type @T@.
\begin{lstlisting}
  case class IRCompiledClo(f: Rep[(ValSt) => ValSt], λ: Lam, ρ: Rep[Env]) extends Def[Value]
  case class IRApClo(fun: Rep[Value], arg: Rep[Value], σ: Rep[Store]) extends Def[ValSt]
\end{lstlisting}

The IR nodes are manipulated by the LMS passes and finally generated as
next-stage Scala programs. To generate code, LMS provides an @emitNode@ method for
programmers to control over what code is generated for each kind of nodes. Here,
we match @IRCompiledClo@ and @IRApClo@, and generate definitions via
@emitValDef@ for them.  For compiled closures, we pass the function @f@ with its
accompanying syntactic term and environment to the next-stage value
representation @CompiledClo@. For applications, we know that @fun@ is
an instance of @CompiledClo@ at the next stage, and it has a callable field
@f@. Hence, we generate a next-stage function application @f(arg, σ)@.  The
invocations of @quote@ are used to generate proper names for the other
next-stage values.
\begin{lstlisting}
override def emitNode(sym: Sym[Any], rhs: Def[Any]) = rhs match {
  case IRCompiledClo(f, λ, ρ) => emitValDef(sym, s"CompiledClo($\textdollar${quote(f)}, $\textdollar${quote(λ)}, $\textdollar${quote(ρ)})")
  case IRApClo(fun, arg, σ) => emitValDef(sym, s"$\textdollar$fun.f($\textdollar${quote(arg)}, $\textdollar${quote(σ)})")
  ...
}
\end{lstlisting}
